Concurrent semantics for left-linear rewriting [in adhesive categories]

### 2024-09-04 domains/domains.tex
section domains

\title{Left-Linear Rewriting in Adhesive Categories}

@@

\begin{abstract}
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in
  parallel. Equivalences identifying rewriting sequences which differ
  only for independent steps are at the core of the theory of
  concurrency of many formalisms.
  % 
  We investigate the issue in the context of the double pushout
  approach to rewriting in the general setting of adhesive categories.
  %, a very general setting for rewriting.  
  While a consolidated theory exists for linear rules, which
  can consume, preserve and generate entities, this paper focuses on
  left-linear rules which may also ``merge'' parts of the state.  This is an apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since % the canonical notion of independence used for linear rewriting
  % systems does not work out of the box, but it requires the
  % restriction to well-behaved classes of left-linear rules.
  % %
  % Moreover,
  a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems, and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

\iffalse
\section*{Summary (with dependencies)}

\subsection*{REWRITING}
\begin{itemize}
\item DPO Rewriting in $\mathcal{M}$-adhesive categories, with left-$\mathcal{M}$-linear rules (from now on shortened in ``linear'')\\
  Observe that rewriting is deterministic (by uniqueness of
  PO-complement Corollary~\ref{lem:radj})

\item Category of abstract derivation for rewriting system $(\X, \R)$
  \begin{itemize}
  \item \emph{DPO-derivation category} $\dpo$: objects are graphs,
    arrows between possibly empty, derivations.
  \item \emph{DPO-abstract decorated derivation category} $\dpi$
  \end{itemize}

\item \emph{Consistent permutations} (Definition~\ref{def:permcon}):
  permutation on the steps of two abstract decorated derivations and
  isomorphism between the colimits such that arrows into the colimits
  ``commutes''.
  \begin{enumerate}
  \item It is unique for consuming
    rewriting systems (Corollary~\ref{cor:unique})
  \item They can be composed
    (Lemma~\ref{lem:sum})
  \end{enumerate}
\item

\item Notions of independence\\ We have several
  possible notions of independence for two direct derivations $\dder{D}
  ; \dder{D'}$
  \begin{enumerate}

  \item \emph{Existence of a filler} There
is a filler for $\dder{D} ; \dder{D'}$ (Definition~\ref{def:filler})

\item \emph{Switchable} There is
  $\dder{D}_1' ; \dder{D}_1$ which is a \emph{switch} of $\dder{D} ;
  \dder{D'}$ (Definition~\ref{de:switch}, being a switch is defined
  axiomatically) [actually, Definition~\ref{de:good-switchable} provides
  a definition of switchable which looks different to me, I would avoid
  this, to be checked]

\item \emph{(Weakly) Sequential
    independent} There is an \emph{independence pair}
  (Definition~\ref{de:sequential-independence}) They are called
  \emph{strongly sequential independent} if the independence pair is
  unique.
\end{enumerate}

We have that $(1) \Rightarrow (2)$ (Theorem~\ref{prop:fil}) and
$(2) \Rightarrow (3)$ while we expect that in general
$(3) \not\Rightarrow (1)$ (but no counterexample). This is instead
known to hold for linear rules.

Some terminology (Definition~\ref{de:good-switchable})
\begin{itemize}

\item \emph{Strong Enforcing rewriting systems}\\ An
  independence pairs induced by a filler is called \emph{strong} A
  rewriting system where all independence pairs are strong is called
  \emph{strong enforcing}

  NOTE: the categories in which
  condition $B$ from~\cite{baldan2011adhesivity} identifies a large
  class of (quasi)adhesive categories where left-linear systems are all
  strong enforcing. Adapted here to M-adhesivity.

\item \emph{Well-Switching rewriting
    systems}\\ Strong Enforcing and existing independence pairs are unique (hence
  switching is unique) (see later for examples)
\end{itemize}

\item \emph{Switch equivalent derivations}
  (concrete)\\
  Definition~\ref{de:switchable}

  THEOREM: Two switch equivalent derivations admit a consistent
  permutation (Lemma~\ref{lem:consperm} ??).

  A consequence of this used for proving that the slice of traces is a
  pre-order.

  The converse holds only when rules are linear [proof clear, but not
  written]  
\end{itemize}

\subsection*{CONCURRENCY}

This part is dealt with for well-switching rewriting systems and depends on this assumption. Note that
\begin{itemize}
\item Linear rewriting systems on adhesive categories are well-switching
\item Left-linear rewriting systems on graph-like categories with generalisation of (i) no isolated nodes in $L$ and (ii) right leg injective on nodes is well-switching
\item Graphs with equivalences (Egg) with left-leg regular mono, right leg mono is well-switching
\end{itemize}

We show
\begin{itemize}

\item Three-steps lemma (Lemma~\ref{lem:indep-global-left}), which is a weakened version of ``independence is global''.
  \begin{itemize}

  \item If in $abc$ I have that $ab$ are independent and $c$ is anticipated, they remain independent.
  \item  If in $abc$ I have that $bc$ are independent and $a$ is moved to the end, they remain independent only if there is an additional condition (essentially requiring that $a$ was not a cause of $c$)
  \end{itemize}

\item No need of useless switches (if two derivations are switch equivalent I can obtain one from another applying only inversions)

\item weak prime domain e connected es semantics
\end{itemize}
\fi

>>>>

\title{Concurrent semantics for left-linear rewriting [in adhesive categories]}

@@

\begin{abstract}
  When moving from linear to left-linear rewriting, many properties
  cease to hold (see concur), and some require to overcome new
  obstacles to be proven. Here, we identify a very general class of
  left-linear rewriting systems on adhesive categories where a
  semantics based on event structures/domains with good properties can
  be defined. Two semantic structures introduced previously, namely
  connected event structures and weakly prime domains, introduced in
  the context of graph rewriting, emerge as canonical structures for
  rewriting with fusion in the context of adhesive categories.
\end{abstract}

@@

\section*{Summary}

\subsection*{1. Background}

CONCUR ruslts: strong enforcing + well-switching => nice behaviour of independence.

\subsection*{2. A class of well behaved rewriting systems}
   [note pag. 11-19]

   Adhesive categories, with all POs, ...,
   coreflection with a subcat from which we take lhs and monos as right legs

   implies

   - strong enforcing + well-switching

   - well-behaved with respect to quotients, e.g.,
   \begin{enumerate}
   \item  if we can rewrite G with L <- K -> R and we take a quotient
     of G which factories through the colimit, then we can still rewrite

   \item if we can rewrite G with L <- K -> R and we take a G' which
     has G as a quotient and we can find a match in G' (which factorises
     through the quotient) then we can still rewrite

   \item existence of the "largest" quotient to which a rule applies

   \item independence "in the colimit" [notes pag. 20-22 for ordinary
     derivations, then adapt to derivations up to fusion, which are ordinary
     derivations in the quotient]
   \end{enumerate}

\subsection*{3. Domain and event structure semantics}

\subsubsection*{3.1 Domain of traces}
finitary, coherent, irreducible algebraic
    characterise irreducibles [traces with last step not switchable]
    [notes page 1-9, holds in general in the setting of concur]

    \subsubsection*{3.2 Weak prime domain and connected event structures}

    - Interchargeability relation and its characterisation via the notion
    of derivation up to fusion

    - derivations up to fusion and properties [notes pag. 24-27]

    - characterising the interchangeability relation using
    derivations up to fusion [some idea at pag. 29, notes]

    - proving the interchangeability axioms to conclude that we
    obtain a weak prime domain [pag. 30-32, notes]
